---
title: "Looking at Pokemon Stats"
author: "Aaqil Zakarya"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(out.width = "100%")
knitr::opts_chunk$set(dpi = 300)

library(readr)
library(ggplot2)
library(dplyr)
library(ggmap)
library(viridis)
library(forcats)
library(smodels)
```


```{r}
pokemon <- read_csv("https://raw.githubusercontent.com/urstats/stat209-s18-AaqilZ/master/Pokemon.csv")
```

## Introduction
For Project C, I gathered the stats of the first - generation (and in my opinion, best) Pokemon. We will primarily be looking at the stats of hp and defense of 100 Pokemons and will see how they compare with the type of the pokemon. Operating under the assumption that as pokemon grow stronger all their stats increase, I expect to see a strong co-relation between the hp and defense stats for each pokemon. Moreover, I wish to see which types of pokemon have higher stats in defense and hp. Over my gaming experience I have noticed that rock type pokemons have better defense stats than most other pokemons. However, I have noticed that no particular type of pokemon dominate in the hp stat so I will put these hypothese to test. 

## Methods

Collection of Data: I collected data on the original 150 pokemon by going on the pokemon website. I left out the types of pokemon that did not have many instances in the original 150 because I wanted to limit the categories I had under the type variable. There are no units to the stats for hp and defense. The numbers assigned to these stats simply are meant to create some form of ordinality among the pokemons. 

I am conducting an analysis on the means of the hp and defense stats for each type of pokemon, with the grass type pokemon as the baseline(for no particular reason) for my comparisions. I am also conducting a linear regression between the hp and defense stats to see the co-relation between the two. 


## Results

```{r}
model <- lm_basic(defense ~ 1 + type, pokemon)

reg_table(model, level = 0.95)
```


```{r}
ggplot(pokemon, aes(type, defense))+
  geom_violin()
```

In this model we can clearly see that the Rock type pokemon dominates in the defense stat. In the violin plot above, we can see that the Rock pokemon has the thickest part of the plot (indicating the mode) much above the modes of any other type of pokemon. This is supported by the estimates provided in the register table.

```{r}
model <- lm_basic(hp ~ 1 + type, pokemon)

reg_table(model, level = 0.95)
```


```{r}
ggplot(pokemon, aes(type, hp))+
  geom_violin()
```

As we can see in the information of the register table, the estimates for the hp stats of non grass type pokemon are not statistically much greater than the estimates for grass type pokemon. The 5% estimates for all the types are in the negative region and the 95% estimates are in the positive region hence we cannot say for sure that any particular type will usually have better hp stats than the grass type. This is represented in the more spread out violin plot above.

```{r}
model <- lm_basic(defense ~ 1 + hp, pokemon)

reg_table(model, level = 0.95)
```


```{r}
qplot(hp, defense, data = pokemon) + 
  geom_smooth(method = "lm")
```

What is quite interesting, and immediately apparent here, is that there does not seem to be a strong co-relation between the hp stats and the defense stats of pokemon at all. This makes one think that the increase in a particular stat does not necessarily translate to other higher stats. Perhaps as Pokemon evolve and grow stronger, their dominant stats increase by much more than other stats do?

## Conclusions

1. I got a result that I had expected. The Rock type pokemon have much higher defense stats than other pokemon. This was supported by the fact that the confidence interval for the Rock type was 32.845  72.99, both of which are large positive values. This suggested that the rock type pokemon would generally have a higher defense stat than the baseline(grass) pokemon. 

2. I got no significant results out of my second analysis. The confidence intervals for all types ranged from negaive to positive numbers, which showed no clear improvement from the baseline stats.

3. I again got no significant results out of my third analysis which was to see if there was a co-relation between the hp and defense stats of pokemon. The confidence interval for hp again ranged from negative to positive and notably the interval had very small values, which indicated that there is very weak co-relation if any. This is quite contrary to what I was expecting to see, which was a strong co-relation between the hp and defense stats. 

## References

https://www.pokemon.com/us/



